### 2025-03-12 paper/preprint/paper.tex
more edits

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{equation}\phantomsection\label{eq-obj}{
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
}\end{equation}

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g., cross-entropy). The two
additional components in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
the different components can be governed by adjusting the strengths of
the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
that CT is highly sensitive to hyperparameter settings but (a) there are
manageable patterns and (b) we can typically identify settings that
improve either plausibility or cost, and commonly both of them at the
same time. Second, we note that the choice of a CE generator has a major
impact on the results. For example, \emph{REVISE} tends to perform the
worst, most likely because it attempts to explain the decisions of a VAE
that is an imperfect surrogate of the original model. Third, increasing
\(T\), the maximum number of steps, generally yields better outcomes
because more CEs can mature in any iteration. Fourth, the impact of
\(\tau\), the required decision threshold is more difficult to predict.
On ``harder'' datasets it may be difficult to satisfy high \(\tau\) for
any given sample (i.e., also factuals) and so increasing this threshold
does not seem to correlate with better outcomes.

\textbf{\emph{Hyperparameters for penalties}} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values for Equation~\ref{eq-obj} penalties, this
effect is not very pronounced.

@@

\textbf{\emph{CT is subject to training instabilities.}} Joint
energy-based models are susceptible to instabilities during training
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
though we depart from the SGLD-based sampling, we still encounter major
variability in the outcomes. CT is exposed to two potential sources of
instabilities: (1) the energy-based contrastive divergence term in
Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
For example, Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023})
recognize this to be a challenge for \emph{ECCCo} and so it may have
downstream impacts on our proposed method. Still, we find that training
instabilities can be largely counteracted with strong regularization, in
particular through \(\lambda_{\text{reg}}\) penalty on the energy.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
discussed in Section~\ref{sec-hyperparameters}, our method requires
careful selection of the hyperparameters. All experiments relied on
grid-search tuning but we recognize that other algorithms may be more
efficient. Notably, counterfactual training is iterative which makes a
variety of methods applicable, including Bayesian (e.g.,
\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
al. 2017}) optimization.

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs (see details in \textbf{?@sec-hardware}). However, there are three
factors that attenuate the impact of this limitation. First, CT provides
counterfactual explanations for the training samples essentially for
free, which may be beneficial in many ADM systems. Second, we find that
CT can retain its value when used as a ``fine-tuning'' training regime
for conventionally-trained models. Third, in principle, CT yields itself
to multi-threaded execution.

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{equation}\phantomsection\label{eq-obj}{
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
}\end{equation}

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between mature
counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularises the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and Mordatch
2020}). The tradeoff between the different components can be governed by
adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).

@@

\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
that CT is highly sensitive to hyperparameter settings but (a) there are
manageable patterns and (b) we can typically identify settings that
improve either plausibility or cost, and commonly both of them at the
same time. Second, we note that the choice of a CE generator has a major
impact on the results. For example, \emph{REVISE} tends to perform the
worst, most likely because it uses a surrogate VAE to generate
counterfactuals which impedes faithfulness
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Third,
increasing \(T\), the maximum number of steps, generally yields better
outcomes because more CEs can mature in each training epoch. Fourth, the
impact of \(\tau\), the required decision threshold is more difficult to
predict. On ``harder'' datasets it may be difficult to satisfy high
\(\tau\) for any given sample (i.e., also factuals) and so increasing
this threshold does not seem to correlate with better outcomes. In fact,
we have generally found that a choice of \(\tau=0.5\) leads to optimal
results because it is associated with high proportions of mature
counterfactuals.

\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
of the energy regularization, \(\lambda_{\text{reg}}\) is highly
impactful; energy must be sufficiently regularized to avoid poor
performance in terms of decreased plausibility and increased costs. The
sensitivity with respect to \(\lambda_{\text{div}}\) and
\(\lambda_{\text{adv}}\) is much less evident. While high values of
\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
combined with high values for any of the other penalties in
Equation~\ref{eq-obj}, this effect is not very pronounced.

@@

\textbf{\emph{CT is subject to training instabilities.}} Joint
energy-based models are susceptible to instabilities during training
(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
though we depart from the SGLD-based sampling, we still encounter major
variability in the outcomes. CT is exposed to two potential sources of
instabilities: (1) the energy-based contrastive divergence term in
Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
For example, Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023})
recognize this to be a challenge for \emph{ECCCo} and so it may have
downstream impacts on our proposed method. Still, we find that training
instabilities can be successfully mitigated by regularizing energy
(\(\lambda_{\text{reg}}\)), generating a sufficiently large number of
counterfactuals during each training epoch and including only mature
counterfactuals for contrastive divergence.

\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
discussed in Section~\ref{sec-hyperparameters}, our method benefits from
tuning certain key hyperparameters. In this work, we have relied
exclusively on grid search for this task. Future work on CT could
benefit from investigating more sophisticated approaches towards
hyperparameter tuning. Notably, counterfactual training is iterative
which makes a variety of methods applicable, including Bayesian (e.g.,
\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
al. 2017}) optimization.

@@

\textbf{\emph{CT increases the training time of models.}} Counterfactual
training promotes explainability through CEs and robustness through AEs
at the cost of longer training times compared to conventional training
regimes. While higher numbers of iterations and counterfactuals per
iteration positively impact the quality of found solutions, they also
increase the required amount of computations. We find that relatively
small grids with 270 settings can take almost four hours for more
demanding datasets on a high-performance computing cluster with 34 2GB
CPUs (see details in \textbf{?@sec-hardware}). However, there are three
factors that attenuate the impact of this limitation. First, CT provides
counterfactual explanations for the training samples essentially for
free, which may be beneficial in many ADM systems. Second, we find that
CT can retain its value when used as a ``fine-tuning'' training regime
for conventionally-trained models. Third, in principle, CT yields itself
to parallel execution, which we have leveraged for our own experiments.

### 2025-03-12 paper/sections/discussion.qmd
more edits

***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method. Still, we find that training instabilities can be largely counteracted with strong regularization, in particular through $\lambda_{\text{reg}}$ penalty on the energy.

***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method requires careful selection of the hyperparameters. All experiments relied on grid-search tuning but we recognize that other algorithms may be more efficient. Notably, counterfactual training is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

@@

***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes. While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations. We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware). However, there are three factors that attenuate the impact of this limitation. First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems. Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models. Third, in principle, CT yields itself to multi-threaded execution.

>>>>

***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes. CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers. For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method. Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch and including only mature counterfactuals for contrastive divergence.

***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method benefits from tuning certain key hyperparameters. In this work, we have relied exclusively on grid search for this task. Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning. Notably, counterfactual training is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

@@

***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes. While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations. We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware). However, there are three factors that attenuate the impact of this limitation. First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems. Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models. Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.
